---
title: "ECON 291 Project"
author: "[First name] [Last name]"
date: "9/2/2022"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

# This is the file provided by the course's instructor, 
# it contains useful libraries and functions
source("ECON 291 Tools.R")

# R libraries are being attached to this project
library(fredr)         # Fed Reserve Bank of St. Louis, FRED
library(bea.R)         # The U.S. Bureau of Economic Analysis
library(tidycensus)    # The U.S. Census Bureau 
library(blsAPI)        # The U.S. Bureau of Labor Statistics

# My access keys, which will be used in this project
key.FRED   = "3f1722e0293651c883a63c15725200f8"
key.BEA    = "5125F980-B756-4FED-9173-7AADF5DE8231"
key.Census = "302ba19863cd8758a59ade8470d22ca410fd5a05"
key.BLS    = "6b732f6c133d48669aadd2790cb71ccc"

# Registering access keys with data providers
fredr_set_key(key.FRED)
census_api_key(key.Census)

```

# The U.S. Economy

## Non-Farm Payroll

Let's 

  (1) retreive the **non-farm payroll** employment data from BLS since 2016 through current year (ID: ```CES0000000001```), 
  (2) then derive the NFP change in the most recent period and 
  (3) graph the whole data set.

```{r echo=FALSE}

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CES0000000001"),   # Non-farm payroll employment
  "startyear"       = 2016,                 # ...first year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )

nfp       <- blsAPI(p,2,TRUE)               # Obtaining the NFP data 
nfp$value <- as.numeric(nfp$value)          # Converting value (thousands of jobs) into numeric
nfp$date  <- as.Date(                       # Creating new variable date to be used in graph 
  paste0("1 ",nfp$periodName," ",nfp$year), 
  format = "%d %B %Y")
nfp <- nfp[order(nfp$date),]                # Ordering data rows by date column
# Most recent NFP increase(+)/decrease(-), unit: thousands of jobs
format(nfp$date[nrow(nfp)],"%B %Y")
diff(nfp$value,1,1)[nrow(nfp)-1]            # The value of the most recent NFP observation

# Plotting the NFP data
ggplot(nfp, aes(x=date,y=value)) + geom_line(col="red") + 
  labs(x="", y="", title="Non-farm payroll employment (thousands)")

```

## Consumer Price Index

Let's grab the most recent inflation data from BLS and graph it. I am interested in 
  
  (1) the general consumer price index, which is CPI-**All items** (ID: ```CUUR0000SA0```),  
  (2) the price index of CPI-**Used cars and trucks** (ID: ```CUUR0000SETA02```),
  (3) and the price index of CPI-**Meats** (ID: ```CUUR0000SAF11211```). 

``` {r echo=FALSE}

# Item codes for CPI details: https://download.bls.gov/pub/time.series/cu/cu.item
# Rule for ID construction: CUUR0000 + Item code
# Example 1: item code for "All Items" is SA0
#            hence, ID for CPI in "All items" is CUUR0000SA0
# Example 2: item code for "Used cars and trucks" is SETA02
#            hence, ID for CPI in "Used cars and trucks" is CUUR0000SETA02

p <- list(                                  # Creating list of data to be retrieved from BLS...
  "seriesid"        = c("CUUR0000SA0",      # All items
                        "CUUR0000SETA02",     # Used cars and trucks
                        "CUUR0000SAF11211"),  # Meats
  "startyear"       = 2016,                 # Start year in the time range,
  "endyear"         = year(Sys.Date()),     # ...current year is the end year of the time range 
  "registrationKey" = key.BLS               # BLS key
  )              

cpi       <- blsAPI(p,2,TRUE)               # Obtaining the CPI data
cpi$value <- as.numeric(cpi$value)
cpi$date  <- as.Date(paste0("1 ",cpi$periodName," ",cpi$year), format = "%d %B %Y")
cpi       <- cpi[order(cpi$date),]

# Adding new column item_name, containing ID names, into the cpi data set
cpi <- bls.cpi.item_name(cpi, p$seriesid)

# Plotting the CPI data as indexes
ggplot(cpi,     aes(x=date,y=value,color=item_name)) + geom_line() + 
  labs(x="", y="", 
       title="CPI-All Items and its two selected components",
       subtitle="Monthly index data")

```

Now, let's derive and plot the year-on-year changes in the CPI indexes:

```{r echo=FALSE}

cpi.yoy <- data.frame()
for (i in 1:length(p$seriesid)) {
    d <- bls.cpi.yoy(cpi,p$seriesid[i])
    cpi.yoy   <- rbind(cpi.yoy, d)
}

# Plotting the CPI data as year-on-year changes
ggplot(cpi.yoy, aes(x=date,y=value,color=item_name)) + geom_line() + 
  labs(x="", y="", 
       title="CPI-All Items and its two selected components",
       subtitle="Year-on-year percentage change in monthly index data")

```
